---
title: "Report on Movie Ratings"
author: "Kevin Camp"
date: '`r Sys.Date()`'
output: pdf_document
header-includes:
    - \usepackage{caption}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1.0 Introduction  

This report satisfies the guidelines for the MovieLens project in the Data Science Capstone course. The  project goal is analyzing a dataset of movie ratings by a collection of users in order to create a recommendation system. The recommendation system will predict ratings by user for movies which that user has not yet rated. The dataset is large; that presents a challenge, in that large datasets are difficult to manage. It also presents an opportunity, in that every data point can be used as a predictor for each missing value in the overall set. The aim of the project is to achieve that using machine learning methods.

# 1.1 Overview  

The project uses a modified version of the MovieLens data with two sets. One set---"edx"---is used to train the model, and contains more than 9 million rows of user--movie ratings. Each row provides columns detailing a user identifier (userId); movie identifier (movieId); movie rating given by the user (rating); time in seconds since the Unix Epoch on midnight, January 1st, 1970 (timestamp); name of the rated movie (title); and genre or genres describing the rated movie (genres). The other set---"validation"---is similar in content to edx, and is used as the final hold-out test set. Table 1.1 presents a selection of rows from edx.  

\captionsetup[table]{labelformat=empty}

```{r create datasets and view edx top, message=FALSE, warning=FALSE, echo=FALSE}
##### Create edx set, validation set (final hold-out test set) #####

# Note: this process could take a couple of minutes

options(repos = list(CRAN="http://cran.rstudio.com/"))

knitr::opts_chunk$set(message = FALSE)

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(kableExtra)) install.packages("kableExtra", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
library(kableExtra)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

download.file("https://www.dropbox.com/s/nspymeso8rmmak1/edx.rds?dl=1", "edx.rds", mode="wb")

download.file("https://www.dropbox.com/s/x0s477b0kzxpl6i/validation.rds?dl=1", "validation.rds", mode="wb")

edx = readRDS("edx.rds")

validation = readRDS("validation.rds")

#view first ten rows
edx %>% slice(1:10) %>% knitr::kable(caption = "Table 1.1. First ten rows of edx dataset",
                                     row.names = FALSE) %>%
  kable_styling(font_size = 10, position = "center",
                latex_options = c("scale_down","HOLD_position"))
```

Additional summary detail on the edx set is included in table 1.2. The table shows the large number of users, movies, and movie genres represented in the data.  

```{r edx summary table, message=FALSE, warning=FALSE, echo=FALSE}

#create summary table
edx_summary <- data.frame(Rows = nrow(edx),
                          Columns = ncol(edx),
                          "Unique users" = n_distinct(edx$userId),
                          "Unique movies" = n_distinct(edx$movieId),
                          "Average rating" = round(mean(edx$rating),2),
                          "Number of genres" = n_distinct(edx$genres),
                          "Date of first rating" = 
                            as.Date(as.POSIXct(min(edx$timestamp), 
                                               origin = "1970-01-01")),
                          "Date of last rating" = 
                            as.Date(as.POSIXct(max(edx$timestamp),
                                               origin = "1970-01-01")),
                          check.names = FALSE)

edx_summary[,1:6] %>% 
  knitr::kable(caption = "Table 1.2. Summary of edx set") %>%
  kable_styling(font_size = 10, position = "center",
                latex_options = c("HOLD_position"))
```

Note the result from multiplying the unique users and unique movies values from table 1.2, 69,878 × 10,677 =  746,087,406, a number larger than the total row count for the edx set (9,000,055). This implies that in the edx data, each user did not rate each movie. It is then useful to view the edx dataset as a matrix, with users represented by rows and movies represented by columns. Chart 1.1 takes a random sample of 100 movies and 100 users and depicts user--movie combinations for which ratings exist with yellow shading. The machine learning algorithm developed in this project will allow us to use values in the entire matrix as predictors for each empty cell.  

\begin{center}
Chart 1.1. Matrix of a random sample of 100 users and 100 movies from edx
\end{center}

```{r matrix sparsity, results='hide', message=FALSE, warning=FALSE, echo=FALSE}
# Take a look at sparsity using a random sample of our matrix of users, movies
install.packages("rafalib")
users <- sample(unique(edx$userId), 100)
rafalib::mypar()
edx %>% filter(userId %in% users) %>% 
  select(userId, movieId, rating) %>%
  mutate(rating = 1) %>%
  spread(movieId, rating) %>% select(sample(ncol(.), 100)) %>% 
  as.matrix() %>% t(.) %>%
  image(1:100, 1:100,. , xlab="Movies", ylab="Users")
abline(h=0:100+0.5, v=0:100+0.5, col = "grey")
```

The edx data contains other useful pieces of information. For example, the timestamp column measures the timing of when users submitted ratings. The rating submission dates in the edx set range from 1995 to 2009, as illustrated in table 1.3.  

```{r edx date range, message=FALSE, warning=FALSE, echo=FALSE}

edx_summary[,7:8] %>% 
  knitr::kable(caption = "Table 1.3. Date range in edx set") %>%
  kable_styling(font_size = 10, position = "center",
                latex_options = c("HOLD_position"))
```

Another example is the genre information for each movie. Exploratory data analysis reveals at least one genre listed for the vast majority of movies recorded in the edx data. In fact, only one movie in the dataset---rated by 7 users---lacks a genre description as demonstrated in table 1.4. Referring back to table 1.1, we see that many movies have multiple genres listed. These variables, along with the others previously mentioned, will be useful predictors in the model.   

```{r edx no genre listed, message=FALSE, warning=FALSE, echo=FALSE}

edx %>%
  filter(genres == "(no genres listed)") %>% 
  knitr::kable(caption = "Table 1.4. Rows in edx lacking genre information",
                                     row.names = FALSE) %>%
  kable_styling(font_size = 10, position = "center",
                latex_options = c("HOLD_position"))
```

# 1.2 Executive summary  

# 2.0 Methods and analysis

# 2.1 Techniques and processes

## 2.1.1 Data cleaning

The timestamp in edx and Tags data set represented seconds since midnight Coordinated Universal Time (UTC) of January 1, 1970. Itw as converted to date for further analysis.

## 2.1.2 Data exploration and visualization

## 2.1.3 Insights gained

# 2.2 Modeling approach

# 3.0 Results

# 3.1 Model results

# 3.2 Model performance

# 4.0 Conclusion

# 4.1 Report summary

# 4.2 Limitations

# 4.3 Future work
